---
title: "3: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Call packages in R

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # Prints 3rd element of vector1 (prints 5.3)

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # creates a 5x4 matrix with numbers from 1 to 20
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # creates 5x4 matrix with numbers in order along rows
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) # creates matrix with row names in spanish and column names in french

matrix1[4, ] # returns the entire 4th row of matrix1
matrix1[ , 3] # returns the entire 3rd column of matrix1
matrix1[c(12, 14)] # return 12th and 14th element of matrix1
matrix1[c(12:14)] # returns 12 through 14th (12th, 13th, 14th) elements of matrix1
matrix1[2:4, 1:3] # returns array of rows 2 through 4 and columns 1 through 3

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]
# Lists can have data of different sizes!

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? Numeric
e <- c("red", "white", "red", NA) # What type of vector? String
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? Logical
dataframe1 <- data.frame(d,e,f) # Data frame composed of the three vectors d, e, and f
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # Names the columns of the df and opens it in a new tab

dataframe1[1:2,] # Prints the first 2 rows in dataframe1 for all columns (ID, color, and passed)
dataframe1[c("ID","Passed")] # Prints columns ID and Passed
dataframe1$ID # Prints ID column

```
Question: How do the different types of data appear in the Environment tab?

> Answer: The vectors and single values are in the "Values" section, data types with multiple columns or sets of data are in "Data".

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: It selects and fills in the first column of the dataframe (ID)

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
highF <- c(44, 47, 51, 39, 37, 43, 45, 48, 56, 51)
lowF <- c(30, 44, 24, 23, 21, 26, 28, 38, 34, 32)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius.

```{r}
highC <- (highF - 32) * 5/9; highC
lowC <- (lowF - 32) * 5/9; lowC
```

Combine your four vectors into a data frame and add informative column names.

```{r}
temps <- data.frame(highF, lowF, highC, lowC)
names(temps) <- c("HighF", "LowF", "HighC", "LowC")
temps
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}
summary(temps) # Summary for all data
sd(temps) # Std. dev. for all data - This doesn't seem to work for the whole data frame
summary(temps$HighF) # Summary for one row
sd(temps$HighF) # Std. dev. for one row
```


## Packages

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your libaray already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a package is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.


```{r}
# We will use the packages dplyr and ggplot2 regularly. 
#install.packages("dplyr") 
#install.packages("ggplot2")
# comment out install commands, use only when needed and re-comment - R Markdown doesnt' knit with install lines active

library(dplyr)
library(ggplot2)

# Some packages are umbrellas under which other packages are loaded
#install.packages("tidyverse")
library(tidyverse)
```

Question: What happens in the console when you load a package?

> Answer: It shows the progress and any associated packages


## Tips and Tricks

* Sequential section headers can be created by using at least four -, =, and # characters.

* The command `require(packagename)` will also load a package, but it will not give any error or warning messages if there is an issue.

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 

* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out. 

* Update your packages regularly! 