---
title: "Assignment 7: GLMs week 2 (Linear Regression and beyond)"
author: "Thomas Hancock"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 25 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory, load the tidyverse, nlme, and piecewiseSEM packages, import the *raw* NTL-LTER raw data file for chemistry/physics, and import the processed litter dataset. You will not work with dates, so no need to format your date columns this time.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1  Load libraries, etc.
getwd() # Check working directory
library(tidyverse)
library(nlme)
library(piecewiseSEM)

# Load data
Litter.Processed <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")
NTL.chemphys.Raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")


#2 Set theme
myTheme <- theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") # Define a theme based off of the classic theme

theme_set(myTheme) # Set defined theme to default

```


## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 

3. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years.
* Only the columns: lakename, year4, daynum, depth, temperature_C
* Only complete cases (i.e., remove NAs)

4. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 

```{r}
#3 Wrangle data 
NTL.chemphys.Filtered <- 
  NTL.chemphys.Raw %>%
  filter(daynum > 181 & daynum < 213) %>% # Filter for days in July
  select(lakename:daynum,depth,temperature_C) %>% # Select specific columns
  na.omit() # Remove entries with any NA values

#4 Use AIC to determine best explanatory variables
NTL.AIC <- lm(data = NTL.chemphys.Filtered, temperature_C ~ year4 + daynum + depth)
step(NTL.AIC) # Compare AIC with explanatory variables removed iteratively

# The AIC was lowest for the model with all explanatory variables
NTL.model1 <- lm(data = NTL.chemphys.Filtered, temperature_C ~ year4 + daynum + depth)
summary(NTL.model1)
```

5. What is the final set of explanatory variables that predict temperature from your multiple regression? How much of the observed variance does this model explain?

> Answer: The set of explanatory variables that best explains temperature variation in the NTL-LTER region is year4, daynum, and depth (all the starting explanatory varaibles). This model explains 74.17% of the variance (multiple linear regression; F-statistic = 9303; df = 3, 9718; R2 = 0.7417).

6. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#6 ANCOVA with interaction effects between lakename and depth
NTL.interaction <- lm(data = NTL.chemphys.Filtered, temperature_C ~ lakename * depth)
summary(NTL.interaction)

```

7. Is there a significant interaction between depth and lakename? How much variance in the temperature observations does this explain?

> Answer: There is a significant interaction between depth and lakename for predicting lake temperature (ANCOVA, p < 0.1 for all levels, R2 = 0.7857). This model accounts for 78.57% of the variance. The interaction between Paul Lake and depth is the least significant of the interactions (p = 0.09664).

8. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#8 Plot of temperature vs depth for each lake
NTL.plot <- 
  ggplot(NTL.chemphys.Filtered, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) + # Set transparency
  geom_smooth(method = "lm", se = FALSE) + # Add best-fit line for each lake (w/out std. error)
  ylim(0,35) + # Restrict y-axis
  labs(x = "Depth, m", y = expression("Temperature " ( degree~C)), color = "") +
  scale_color_viridis_d()

print(NTL.plot)
  

```

9. Run a mixed effects model to predict dry mass of litter. We already know that nlcdClass and functionalGroup have a significant interaction, so we will specify those two variables as fixed effects with an interaction. We also know that litter mass varies across plot ID, but we are less interested in the actual effect of the plot itself but rather in accounting for the variance among plots. Plot ID will be our random effect.

a. Build and run a mixed effects model.
b. Check the difference between the marginal and conditional R2 of the model. 

```{r}
# 9a Mixed effects model
LitterTest.mixed <- lme(data = Litter.Processed, # Mixed effects model
                        dryMass ~ nlcdClass * functionalGroup, # Fixed effects
                        random = ~1 | plotID) # Random effects
summary(LitterTest.mixed)

# 9b Compare marginal (no random effect) and conditional (mixed effects) r-squared values
rsquared(LitterTest.mixed) # Show R^2 value

```

b. continued... How much more variance is explained by adding the random effect to the model? 

> Answer: 2.1% more variance is explained by the random effects model compared to just the fixed effects (R2_conditional = 0.268, R2_marginal = 0.247).

c. Run the same model without the random effect.
d. Run an anova on the two tests. 
```{r}
# 9c Fixed effects model
LitterTest.fixed <- gls(data = Litter.Processed, dryMass ~ nlcdClass * functionalGroup)
summary(LitterTest.fixed)

# 9d Compare results using ANOVA
anova(LitterTest.mixed, LitterTest.fixed)

```

d. continued... Is the mixed effects model a better model than the fixed effects model? How do you know?

> Answer: The models have a significantly different fit (ANOVA, p < 0.0001), and the mixed effects model has a lower AIC (9038 vs 9058), so it is the better model.