---
title: "Assignment 2: Coding Basics"
author: "Tommy Hancock"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
num_list <- seq(1, 100, 4) # Generate a sequence of numbers from 1 to 100 by 4s
#2. 
mean(num_list) # Calculate the mean
median(num_list) # Calculate the median
#3. 
mean(num_list) > median(num_list) # Return TRUE if the mean is greater than the median
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
names <- c("Scotty", "Lily", "Samantha", "Ronald") # String vector of names
scores <- c(76, 98, 86, 82) # Numeric vector with test scores
pass <- scores >= 50 # Logical vector determining whether the student passed

test_results <- data.frame(names, scores, pass) # Create a dataframe from the vectors
names(test_results) <- c("Name", "Score", "Pass")

test_results
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: The different columns of the dataframe have different types of data in them (string, numeric, and logical), whereas a matrix requires all the data to be of the same type.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
pass_func1 <- function(x) { # Function using if/else to determine if a score passes
 if (x >= 50) {
   print("Pass")
 } 
  else {
    print("Fail")
  }
}

pass_func2 <- function(x) { # Function using ifelse to determine if a score passes
  ifelse(x >= 50, "Pass", "Fail")
}

pass_func1(scores)
pass_func2(scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The ifelse function worked when applied to the "scores" vector, whereas the if/else function did not work for the vector. (Both worked for individual scores, though.) I got these results because the if/else statement requires a single value to evaluate, whereas the ifelse function is applied sequentially to all values within the vector.
